---
title: "Incremental Graphs"
author: "Wei Yang Tham"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Intro

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
  comment = NA, dpi = 300, echo = FALSE, cache = F,
  fig.align = "center", out.width = "100%", fig.asp = 0.618)
```

- Adding elements of a chart incrementally can help immensely with clarity
- These slides contain a few pointers for making an incremental chart with `ggplot2`
- tl;dr - use `geom_blank`

---

# Set up

We'll use a subset of the `gapminder` dataset and a little bit of data manipulation with `tidyverse` packages. 

```{r, echo=TRUE, tidy=FALSE}
# install.packages("tidyverse")
# install.packages("gapminder")

library(tidyverse)
library(gapminder)

theme_set(ggthemes::theme_fivethirtyeight(base_size = 18))

minigap = gapminder %>%
  filter(country %in% c("Bolivia", "Rwanda", "Germany")) %>% 
  mutate(country = 
           fct_relevel(country, c("Rwanda", "Bolivia", "Germany")))

```

---

# Target plot

```{r}
original = minigap %>% 
  ggplot(aes(year, gdpPercap, colour = country)) + 
  geom_line(size = 1.2) + 
  scale_colour_manual(values = c("red", "blue", "green")) 

original
```

---

# Goal

- Show Rwanda first, then Bolivia, then Germany
- Natural first step is to tell `geom_line` to only plot Rwanda

```{r, eval=FALSE, echo = TRUE}
minigap %>% 
  ggplot(aes(year, gdpPercap, colour = country)) + 
  geom_line(data = filter(minigap, country == "Rwanda")) + 
  scale_colour_manual(values = c("red", "blue", "green")) 
```

---

# First try

```{r}
minigap %>% 
  ggplot(aes(year, gdpPercap, colour = country)) + 
  geom_line(data = filter(minigap, country == "Rwanda")) + 
  scale_colour_manual(values = c("red", "blue", "green")) + 
  labs(title = "Y-axis range has changed!", 
       subtitle = "Legend only has one country")
```


---

# geom_blank does the trick

- `geom_blank` preserves the axis scales

```{r, eval = F, echo = T}
p = ggplot(minigap, aes(year, gdpPercap, colour = country)) + 
  geom_blank() + 
  scale_colour_manual(values = c("red", "blue", "green")) + 
  NULL
```

- From here,  we can add each country's line layer by layer, starting from

```{r, eval=FALSE, echo=TRUE}

p + 
  geom_line(data = filter(minigap, country == "Rwanda")) 
```

and ending with 

```{r, eval = FALSE, echo=TRUE}
# Final plot
p + 
  geom_line(data = filter(minigap, country == "Rwanda"), size = 1.2) + 
  geom_line(data = filter(minigap, country == "Bolivia"), size = 1.2) + 
  geom_line(data = filter(minigap, country == "Germany"), size = 1.2)
```


---

# Rwanda

```{r}

p = ggplot(minigap, aes(year, gdpPercap, colour = country)) + 
  geom_blank() + 
  scale_colour_manual(values = c("red", "blue", "green")) + 
  NULL

p + geom_line(data = filter(minigap, country == "Rwanda"), size = 1.2) 

```

---

# Rwanda, Bolivia

```{r}
p + 
  geom_line(data = filter(minigap, country == "Rwanda"), size = 1.2) + 
  geom_line(data = filter(minigap, country == "Bolivia"), size = 1.2)

```

---

# Rwanda, Bolivia, and Germany

```{r}
p + 
  geom_line(data = filter(minigap, country == "Rwanda"), size = 1.2) + 
  geom_line(data = filter(minigap, country == "Bolivia"), size = 1.2) + 
  geom_line(data = filter(minigap, country == "Germany"), size = 1.2)
```

---

# Additional notes

- Some people even suggest building the graph starting from the axes, without any data plotted i.e. start with

```{r, out.width='60%'}
ggplot(minigap, aes(year, gdpPercap, colour = country)) + 
  geom_blank() + 
  scale_colour_manual(values = c("red", "blue", "green")) + 
  labs(title = "There is no legend")
```

- Unfortunately I don't know how to do this and still have a legend **(if you do, let me know!)**
- Generally, I prefer to do without a legend and have labels directly on the graph. If you do that, then this issue doesn't matter

---

# Sources + Resources

- [Stack Overflow](https://stackoverflow.com/questions/22276761/how-to-add-a-legend-label-in-a-plot-when-there-is-no-data-points-with-that-labe)

- [Better Presentations by Jon Schwabish](https://policyviz.com/better-presentations/)

- [Paul Goldsmith-Pinkham's beamer tips](https://github.com/paulgp/beamer-tips)